Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-05-25 TEX/The autocorrelation functions in cryoEM.tex
Latest comments of Amit for intro, and of Will for appendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\numberwithin{equation}{section}
%\numberwithin{figure}{section}
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}[section]
\theoremstyle{definition}
\newtheorem{defn}[thm]{\protect\definitionname}
\theoremstyle{remark}
\newtheorem{claim}[thm]{\protect\claimname}
\theoremstyle{plain}
\newtheorem{lem}[thm]{\protect\lemmaname}

@@

The 3-D Fourier transform of an L-bandlimited 3-D volume (e.g., particle) can be expanded by  spherical harmonics:
\begin{equation} \label{eq:volume}
\hat{V}(k,\theta,\phi)  = \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k)Y_\ell^{m}(\theta,\phi),
\end{equation}
where $(\theta,\phi)$ are two angles on the sphere, $k$ is the radial coordinate, $Y_{\ell}^m(\theta,\phi)$ is the spherical harmonic of degree $\ell$ and order $m$ and  $A_{\ell,m}(k)$ are the associated spherical harmonics coefficients, to be estimated. 
A rotation of the volume by $\omega\in SO(3)$ can be described using the Wigner D-function $D_{m,m'}^{\ell}$:
\begin{equation}
\begin{split}
(R_\omega \hat{V})(k,\theta,\phi)  &= \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k)(R_\omega Y_\ell^{m})(\theta,\phi) \\
&= \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k) 
\sum_{m'=-\ell}^{\ell}{D_{m,m'}^{\ell}(\omega)}Y_\ell^{m'}(\theta,\phi).
\end{split}
\end{equation}

By the Fourier slice theorem, each cryo--EM measurement (projection) is equivalent (through 2-D Fourier transform) to the slice of $\hat{V}$, associated with $\theta= \pi/2$, after $\hat{V}$ was rotated by $\omega\in SO(3)$. Explicitly, the Fourier transform of a projection from the viewing direction $\omega$ is related to the spherical harmonic coefficients of the object through:
\begin{equation} \label{eq:P}
\begin{split}
P_\omega(k,\phi)  = \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k) 
\sum_{m'=-\ell}^{\ell}{D_{m,m'}^{\ell}(\omega)}Y_\ell^{m'}(\pi/2,\phi).
\end{split}
\end{equation}

Next, we want to relate the projections $P_\omega$ with the mean and the autocorrelation functions computed directly from the micrograph.
The mean of the micrograph is proptional to 
\begin{equation}
M_1 \propto \sum_{n=1}^N \sum_{x,y}P_n(x,y),
\end{equation}
where $P_n$ denotes the $n$th projection. By taking $n\to\infty$, we get  
\begin{equation}
M_1 \propto \sum_{x,y}\int_{\omega}P_\omega(x,y)\rho(\omega) d\omega,
\end{equation}
where $\rho(\omega)$ denotes the (possibly unknown) viewing direction distribution over $SO(3)$.

@@

In order to determine the particle, by~\eqref{eq:volume} one needs to estimate order of $L^3$ spherical harmonics coefficients. If the pixel size is proportional to $1/L$ (to match the volume's resolution), then $M_3$ provides order of $L^4$ equations involving triple products of $P_\omega$. 
However, 
since  the in-plane rotation of each particle image is usually uniformly distributed, $M_3$  depends on only three parameters: the length of the vector $(\Delta_x^1,\Delta_y^1)$, the length of the vector  $(\Delta_x^2,\Delta_y^2)$ and the  angle between the two vectors.  Therefore,  $M_3$  provides only $\sim L^3$ equations. 
 Since $P_\omega$ depends (after coordinate transformation) linearly in the spherical harmonic coefficients through~\eqref{eq:P}, this means we have a system of $\sim L^3$ cubic equations in the $\sim L^3$ sought parameters. Importantly, the coefficients of these equations can be estimated from the micrographs directly, without particle picking stage.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
%\numberwithin{equation}{section}
%\numberwithin{figure}{section}
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}
\theoremstyle{definition}
\newtheorem{defn}[thm]{\protect\definitionname}
\theoremstyle{remark}
\newtheorem{claim}[thm]{\protect\claimname}
\theoremstyle{plain}
\newtheorem{lem}[thm]{\protect\lemmaname}

@@

The 3-D Fourier transform of an L-bandlimited 3-D volume (e.g., particle) can be expanded into spherical harmonics:
\begin{equation} \label{eq:volume}
\hat{V}(k,\theta,\phi)  = \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k)Y_\ell^{m}(\theta,\phi),
\end{equation}
where $\theta \in [0,\pi)$ is the polar angle, $\phi \in [0,2\pi)$ is the azimuthal angle, $k$ is the radial coordinate, $Y_{\ell}^m(\theta,\phi)$ is the spherical harmonic of degree $\ell$ and order $m$ and  $A_{\ell,m}(k)$ are the associated spherical harmonics coefficients. The goal is to estimate the functions $A_{l,m}$.
A rotation of the volume by $\omega\in SO(3)$ can be described using the Wigner D-function $D_{m,m'}^{\ell}$:
\begin{equation}
\begin{split}
(R_\omega \hat{V})(k,\theta,\phi)  &= \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k)(R_\omega Y_\ell^{m})(\theta,\phi) \\
&= \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k) 
\sum_{m'=-\ell}^{\ell}{D_{m,m'}^{\ell}(\omega)}Y_\ell^{m'}(\theta,\phi).
\end{split}
\end{equation}

By the Fourier slice theorem, the Fourier transform of each cryo-EM measurement (that is, each projection) is a slice of $\hat V$, associated with $\theta= \pi/2$, after $\hat{V}$ was rotated by $\omega\in SO(3)$. Explicitly, the Fourier transform of a projection from the viewing direction $\omega$ is related to the spherical harmonic coefficients of the object through:
\begin{equation} \label{eq:P}
\begin{split}
\hat P_\omega(k,\phi)  = \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k) 
\sum_{m'=-\ell}^{\ell}{D_{m,m'}^{\ell}(\omega)}Y_\ell^{m'}(\pi/2,\phi).
\end{split}
\end{equation}

Next, we relate the projections $P_\omega$ to the mean and the autocorrelation functions, both computable from the observed micrographs.
The mean of the micrograph is proportional to 
\begin{equation}
M_1 \propto \sum_{n=1}^N \sum_{x,y}P_{\omega_n}(x,y),
\end{equation}
where $\omega_n$ denotes the viewing direction of the $n$th projection. By taking $n\to\infty$, we get  
\begin{equation}
M_1 \propto \sum_{x,y}\int_{\omega}P_\omega(x,y)\rho(\omega) d\omega,
\end{equation}
where $\rho(\omega)$ denotes the (possibly unknown) viewing direction distribution over $SO(3)$.

@@

In order to determine the particle, by~\eqref{eq:volume} one needs to estimate order of $L^3$ spherical harmonics coefficients. If the pixel size is proportional to $1/L$ (to match the volume's resolution), then $M_3$ provides order of $L^4$ equations involving triple products of $P_\omega$. 
However, 
since  the in-plane rotation of each particle image is usually uniformly distributed, $M_3$  depends on only three parameters: the length of the vector $(\Delta_x^1,\Delta_y^1)$, the length of the vector  $(\Delta_x^2,\Delta_y^2)$ and the  angle between the two vectors.  Therefore,  $M_3$  provides only $\sim L^3$ equations. 
 Since $P_\omega$ depends (after coordinate transformation) linearly on the spherical harmonic coefficients through~\eqref{eq:P}, this means we have a system of $\sim L^3$ cubic equations in the $\sim L^3$ sought parameters. Importantly, the coefficients of these equations can be estimated from the micrographs directly, without particle picking.

### 2018-05-25 TEX/draft4.tex
Latest comments of Amit for intro, and of Will for appendix

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical $\SNR$. This fact has been recognized early on by the cryo-EM community. In particular, in a highly cited paper from 1995, Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{quote}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{quote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight.
%This has often been understood as meaning that particle reconstruction itself is impossible for such molecules~\cite{khoshouei2017cryo,liu2018nearatomic,liang2017phase}. \TODO{This list may be unfair, as some of these papers (in particular, \cite{liang2017phase}) just state that small molecules are very challenging, which is true.}
As a result, it is impossible to reconstruct such small molecules by any of the existing computational pipelines for single particle analysis, as the particles themselves cannot be picked from the micrographs.
%
%\TODO{either here, or in a close neighborhood of this paragraph you should mention the advances that the field has made in recent years towards reconstruction of small size molecules: mention Volta phase plate and the scaffolding cage (and cite the papers I sent you).}
This has motivated recent technical advances in the field, including the use of Volta phase plates~\cite{khoshouei2017cryo,liang2017phase} and scaffolding cages~\cite{liu2018nearatomic}.
%
%
Despite this progress, detecting small molecules in the micrographs remains a challenge.

@@

%\TODO{Paragraph about the simplified model}
As a proof of concept \TODO{To support our point?}, we study a simplified model where an unknown image appears numerous times at unknown locations in several micrographs, each affected by additive Gaussian noise---see Figure~\ref{fig:micro_example} for an illustration.
The goal is to estimate the planted image. The task is interesting in particular when the $\SNR$ is low enough that particle picking cannot be done reliably.
%Figure~\ref{fig:Einst_example} displays the results of our method, detailed in Section~\ref{sec:results}.
A precise mathematical formulation of the model, including an extension where more than one planted images are to be recovered, is provided in Section~\ref{sec:methods}.
To be clear, we do not consider here many prominent features of real SPR experiments and do not aim to reconstruct any 3-D structure. Instead, we solve a simpler problem that we believe captures key elements of the SPR problem. We note that similar models emerge in spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.

@@

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{shatsky2009method,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using cross-correlation and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates are required.

\TODO{What about this one? \cite{vanheel1992correlation}}

>>>>

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical $\SNR$. This fact has been recognized early on by the cryo-EM community. In particular, in a highly cited paper from 1995, Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{quote}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{quote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight.
%This has often been understood as meaning that particle reconstruction itself is impossible for such molecules~\cite{khoshouei2017cryo,liu2018nearatomic,liang2017phase}. \TODO{This list may be unfair, as some of these papers (in particular, \cite{liang2017phase}) just state that small molecules are very challenging, which is true.}
As a result, it is impossible to reconstruct such small molecules by any of the existing computational pipelines for single particle analysis, as the particles themselves cannot be picked from the micrographs.
%
%\TODO{either here, or in a close neighborhood of this paragraph you should mention the advances that the field has made in recent years towards reconstruction of small size molecules: mention Volta phase plate and the scaffolding cage (and cite the papers I sent you).}
This has motivated recent technical advances in the field, including the use of Volta phase plates~\cite{khoshouei2017cryo,liang2017phase} and scaffolding cages~\cite{liu2018nearatomic}.
%
%
Despite this progress, detecting small molecules (below 50 kDa) in the micrographs remains a challenge.

@@

%\TODO{Paragraph about the simplified model}
As a proof of concept, we study a simplified model where an unknown image appears numerous times at unknown locations in several micrographs, each affected by additive Gaussian noise---see Figure~\ref{fig:micro_example} for an illustration.
The goal is to estimate the planted image. The task is interesting in particular when the $\SNR$ is low enough that particle picking cannot be done reliably.
%Figure~\ref{fig:Einst_example} displays the results of our method, detailed in Section~\ref{sec:results}.
A precise mathematical formulation of the model, including an extension where more than one planted images are to be recovered, is provided in Section~\ref{sec:methods}.
To be clear, we do not consider here many prominent features of real SPR experiments and do not aim to reconstruct any 3-D structure. Instead, we solve a simpler problem that we believe captures key elements of the SPR problem. We note that similar models emerge in spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.

@@

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{vanheel1992correlation,shatsky2009method,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using cross-correlation and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates are required.
